---
title: "DATA608 - Assignment 1"
author: "Mael Illien"
date: "9/5/2020"
output:
  html_document: 
    highlight: pygments
    theme: cerulean
    toc: yes
    toc_float: yes
  html_notebook: default
---

**Principles of Data Visualization and Introduction to ggplot2**

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

##### I have provided you with data about the 5,000 fastest growing companies in the US, as compiled by Inc. magazine. lets read this in:

```{r}
inc <- read.csv("https://raw.githubusercontent.com/charleyferrari/CUNY_DATA_608/master/module1/Data/inc5000_data.csv", header= TRUE)
```

And lets preview this data:

```{r}
head(inc)
```
```{r}
summary(inc)
```

##### Think a bit on what these summaries mean. Use the space below to add some more relevant non-visual exploratory information you think helps you understand this data:

The summaries below help to understand the skew of the data the via the total, average and mean columns and also provide an aggregate of the revnue per state and per industry. 

```{r echo=FALSE, message=FALSE, warning=FALSE}
#inc$Employees[is.na(inc$Employees)] <- 0 # Replace NAs with in
inc <- na.omit(inc)

state_data <- inc %>% group_by(State) %>% 
      summarise(CompanyCount = n(), TotEmployee = sum(Employees), AvgEmployee = round(mean(Employees),2), StdEmployee = round(sd(Employees),2), SumRevenue_b = sum(Revenue)/(10^9)) %>% 
      arrange(desc(CompanyCount,na.rm=TRUE)) %>% top_n(10)
knitr::kable(state_data, caption = "Summary: Top 10 State by Company Count")
```

```{r echo=FALSE}
industry_data <- inc %>% group_by(Industry) %>% 
      summarise(CompanyCount = n(), TotEmployee = sum(Employees), AvgEmployee = round(mean(Employees),2), StdEmployee = round(sd(Employees),2), SumRevenue_b = sum(Revenue)/(10^9)) %>% 
  arrange(desc(CompanyCount,na.rm=TRUE))
knitr::kable(industry_data, caption = "Summary: Industries")
```




## Question 1

##### Create a graph that shows the distribution of companies in the dataset by State (ie how many are in each state). There are a lot of States, so consider which axis you should use. This visualization is ultimately going to be consumed on a 'portrait' oriented screen (ie taller than wide), which should further guide your layout choices.

I chose to use a sorted bar graph. The large number of states justifies the flip in axes. The bar graph uses length to display information which is is visually easy to interpret, while the sorting eliminates having to visually compare non-adjacent bars. 

```{r echo=FALSE}
#inc$Employees[is.na(inc$Employees)] <- 0 # Replace NAs with 0
comp_count <- inc %>% 
  group_by(State) %>% 
  summarise(CompanyCount = n()) %>% 
  arrange(desc(CompanyCount,na.rm=TRUE))
```

```{r fig1, echo=FALSE, fig.height=7, fig.width=7, message=FALSE, warning=FALSE}
ggplot(data = comp_count, aes(x=reorder(State, CompanyCount), y=CompanyCount)) + 
  geom_bar(stat = 'identity', fill= 'lightblue', width = 0.8) + 
  geom_text(aes(label = CompanyCount), hjust = -0.5, color='black') +
  coord_flip() + 
  ggtitle("Company Distribution by State") +
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        axis.text.x = element_blank(),
        panel.background = element_blank())
```

## Question 2

##### Lets dig in on the state with the 3rd most companies in the data set. Imagine you work for the state and are interested in how many people are employed by companies in different industries. Create a plot that shows the average and/or median employment by industry for companies in this state (only use cases with full data, use R's `complete.cases()` function.) In addition to this, your graph should show how variable the ranges are, and you should deal with outliers.

I chose the boxplot and the log transform because the it had the nicest spread of data over the plot and allowed for the larger industries not to visually swamp out the smaller ones. Sorting the boxplot by median also helps to makes the pattern in the mean more discernable. While the data may have a nice spread, the reader may have difficulties interpreting the mean and median values relative to the log scale.   

```{r include=FALSE}
ny_state <- inc %>% filter(complete.cases(inc)) %>% filter(State == 'NY') %>% arrange(desc(Employees))
```

```{r eval=FALSE, include=FALSE}
ggplot(data = ny_state, aes(x=Industry, y=Employees)) + 
  geom_boxplot(outlier.colour = "red", outlier.shape = 1) +
  stat_summary(fun=mean, geom="point", shape=20, size=2, color="red", fill="red") +
  scale_y_continuous(trans='log10') +
  coord_flip() #+
  #theme(panel.background = element_blank())
```

```{r fig2, echo=FALSE}
ggplot(data = ny_state, aes(x=reorder(Industry, Employees, FUN=median), y=Employees)) + 
  geom_boxplot(outlier.colour = "red", outlier.shape = NA) +
  stat_summary(fun.y=mean, geom="point", shape=20, size=2, color="blue", fill="red", show.legend=TRUE) +
  scale_y_continuous(trans='log10', breaks=c(1,5,10,50,100,500,1000,5000)) +
  coord_flip() +
  ggtitle("Distribution of Employee Counts in NY State Industries") +
  theme(panel.background = element_blank(),
        panel.grid.major.x = element_line(colour = "gray"),
        axis.title.y = element_blank(),
        axis.title.x = element_blank()) +
  annotate(geom="text", x=20, y=13000, label="mean",color="black") +
  geom_point(aes(x=20, y=6500), size=1, color="blue")
```

```{r fig2a, eval=FALSE, include=FALSE}
ggplot(data = ny_state, aes(x=reorder(Industry, Employees, FUN=median), y=Employees)) + 
  geom_boxplot(outlier.colour = "red", outlier.shape = NA) +
  stat_summary(fun.y=mean, geom="point", shape=20, size=2, color="blue", fill="red") +
  #coord_cartesian(ylim = c(0, 5000)) +
  scale_y_continuous(breaks=seq(0,5000,250), limits=c(0,1000)) +
  coord_flip()
```

## Question 3

##### Now imagine you work for an investor and want to see which industries generate the most revenue per employee. Create a chart that makes this information clear. Once again, the distribution per industry should be shown.

Here, a rotated and sorted bar seemed again to be the most appropriate choice. The bar labels were divided by 10^3 and rounded to make the number more digestible to the user. 

```{r include=FALSE}
ind_rev_per_em <- inc %>% filter(complete.cases(inc)) %>% 
  group_by(Industry) %>% 
  summarise(TotalRevenue = sum(Revenue), TotalEmp = sum(Employees)) %>% 
  mutate(RevPerEmp = TotalRevenue/TotalEmp) %>% 
  arrange(desc(RevPerEmp))
ind_rev_per_em
```

```{r fig3, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = ind_rev_per_em, aes(x=reorder(Industry, RevPerEmp), y=RevPerEmp)) + 
  geom_bar(stat = 'identity', fill= 'lightblue', width = 0.8) + 
  geom_text(aes(label = paste0("$", round(RevPerEmp/1000,0), 'k')), hjust = -0.1, color='black') +
  #aes(label=ifelse(EUR>100,paste0("€", round(EUR,0)),""),)
  scale_y_continuous(labels=scales::dollar_format()) +
  coord_flip() + 
  ggtitle("Industry Revenue per Employee") +
  theme(
        axis.ticks = element_blank(),
        axis.text.x = element_blank(),
        axis.title = element_blank(),
        panel.background = element_blank()) +
  expand_limits(y = c(0, 1500000))

```

